Is high social class always beneficial for survival? Northern Sweden 1801--2013

### 2018-02-04 article3b.Rmd
introduce median life

\begin{enumerate}
\item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
\item {\tt middle}, Hisclass 3, 4, 5, 6 and 8. Lower managers,
farmers and lower professionals, clerical and sales personell.
\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
farm workers.
\end{enumerate}

@@

```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '50%', fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

The analyses are performed in the **R** environment for statistical computing
and graphics [@cran], especially using the package **eha**
[@gb12;@eha]. In the analysis, data are divided into two age groups,
40--64 and 65--89 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
but we then combine results in order to show the development of the social patterning of mortality during the studied period.

@@

The models are fitted separately for sex and each period, from  1801 to
2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
**civst** (marital status). The cumulative
hazards at the end of the age period (65 if 40--64, 90 if 65--89), conditional on survival 
to the start of the age period, are the
main target in our analysis. These numbers are used in the process of
graphically illustrating the development over time.

@@

```{r colorc}
## Colors for plotting strata:
scol <- c("blue", rep("darkgreen", 2), "red")
## Common for all groups:

@@

## Ages 40--64

```{r getske4060}
cap1w <-  "Cumulative hazards for social class by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##

cap0 <- "Cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
scap0 <- "Cumulative hazard of dying before age 65 for a 40-year old by social class."
cap1 <- "Cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
scap1 <- "Cumulative hazard of dying before age 90 for a 65-year old by social class."
cap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
scap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class."
cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
scap12log <- "Probability by social class, ages 40--64, last years."

cap2w <-  "Cumulative hazards for social class by time period, ages 65--89, women."
scap2w <-  "Cumulative hazards, ages 65--89, women."
cap2m <-  "Cumulative hazards for social class by time period, ages 65--89, men."
scap2m <-  "Cumulative hazards, ages 65--89, men."
##
cap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
scap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class."
cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
scap22log <- "Probability by social class, ages 65--89, last years."

##
vb4064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
vb4064$age <- factor(vb4064$age)
##vb4064$hisclass <- vb4064$hisclass40
```

```{r hazcalw}
source("R/getPHglm2.R")
source("R/getPHglm.R")
##nper <- 4
nper <- length(labb)
resw <- getPHglm2(vb4064[vb4064$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb4064[vb4064$sex == "male", ],
            nper, labb)
```

```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Women 40-64", ylim = c(0, 1))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Men 40-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswtab,cap="Table"}
vb4064$social_class <- vb4064$hisclass
vb4064$marital_status <- vb4064$civst
vb4064$urban_or_rural <- vb4064$urban
vb4064$age_group <- vb4064$age
fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb4064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
dr
##knitr::kable(dr)
##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 40-64")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 40-64")##, exclude = 5:6)
par(oldpar)
```

@@

## Ages 65--89

```{r getske6089}
vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
##vb6589$hisclass <- vb6589$hisclass60
```

We now turn to the elderly, those who in most cases have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.

```{r cumswtabold,cap="Table"}
vb6589$marital_status <- vb6589$civst
vb6589$social_class <- vb6589$hisclass
vb6589$urban_or_rural <- vb6589$urban
vb6589$age_group <- vb6589$age
fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb6589, family = poisson)
drold <- drop1(fittold, test = "Chisq")
##knitr::kable(dr)
drold
##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--89."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
##source("R/plotED.R")
source("R/getPHglm.R")
resw <- getPHglm2(vb6589[vb6589$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb6589[vb6589$sex == "male", ],
            nper, labb)

oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = FALSE,
                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-89")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = FALSE,
                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-89")##, exclude = 5:6)
par(oldpar)
```

```{r cumsw2, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE, 
                    ylab = "H(90|65)",
                main = "Women 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(90|65)",
                main = "Men 65-89", ylim = c(0, 0.6))##, exclude = 5:6)
par(oldpar)
```

```{r cumswoldtab,results='asis',eval=FALSE}
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
            data = vb6589, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-89.")
```

>>>>

\begin{enumerate}
\item {\tt elite}, Hisclass 1 and 2. Higher managers and higher professionals.
\item {\tt middle}, Hisclass 3, 4, 5, and 6. Lower managers,
farm hands and lower professionals, clerical and sales personell.
\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
farm workers.
\item {\tt farmer}, Hisclass 8. Farmers. 
\end{enumerate}

@@

```{r hiscskumm, fig.cap = "Exposure by social class (from the top: elite, middle, workers, farmers, missing) and time period, men.", fig.scap="Exposure by social class and time period, men.", out.height = '50%', fig.height=4}
xx <- with(skel[skel$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "darkgreen", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

```{r hiscskumw, fig.cap = "Exposure by social class (from the top: elite, middle, workers, missing) and time period, women.", out.height = '50%', fig.scap="Exposure by social class and time period, women.",fig.height=4}
xx <- with(skel[skel$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xy <- with(skume[skume$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xz <- with(foli[foli$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
xxx <- rbind(xx, xy, xz)
yy <- round(prop.table(xxx, 1) * 100, 1)
barplot(t(yy)[5:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
```

@@

The analyses are performed in the **R** environment for statistical computing
and graphics [@cran], especially using the package **eha**
[@gb12;@eha]. In the analysis, data are divided into two age groups,
30--64 and 65--99 years of age. The model controls for marital status and whether the individual resides in the urban or rural parts of the regions. Each period is analysed separately, 
but we then combine results in order to show the development of the social patterning of mortality during the studied period.

@@

The models are fitted separately for sex and each period, from  1801 to
2013. We stratify on social class and include the covariates **urban** (TRUE or FALSE) and
**civst** (marital status). The *cumulative hazards* at the end of the age period *30--64* 
conditional on survival to age 30,  and  *remaining expected life at 65* are the
main targets in our analysis. These numbers are used in the process of
graphically illustrating the development over time.

@@

```{r colorc}
## Colors for plotting strata:
scol <- c("black", "green", "blue", "darkgreen", "red")
## Common for all groups:

@@

## Ages 30--64

```{r getske3064}
cap1w <-  "Cumulative hazards for social class by time period, ages 30--64, women."
scap1w <-  "Cumulative hazards, ages 30--64, women."
cap1m <-  "Cumulative hazards for social class by time period, ages 30--64, men."
scap1m <-  "Cumulative hazards, ages 30--64, men."
##

cap0 <- "Cumulative hazard of dying before age 65 for a 30-year-old by social class and period, women (top) and men (bottom)."
scap0 <- "Cumulative hazard of dying before age 65 for a 30-year old by social class."
cap1 <- "Remaining years of life for a 65-year-old by social class and period, women (top) and men (bottom)."
scap1 <- "Remaining life for a 65-year old by social class."
cap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom)."
scap12 <- "Relative cumulative hazard of dying before age 65 for a 40-year-old by social class."
cap12log <- "Probability of dying before age 65 for a 40-year-old by social class and period, women (top) and men (bottom). The last decades."
scap12log <- "Probability by social class, ages 40--64, last years."

cap2w <-  "Cumulative hazards for social class by time period, ages 65--99, women."
scap2w <-  "Cumulative hazards, ages 65--99, women."
cap2m <-  "Cumulative hazards for social class by time period, ages 65--99, men."
scap2m <-  "Cumulative hazards, ages 65--99, men."
##
cap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom)."
scap22 <- "Relative cumulative hazard of dying before age 90 for a 65-year-old by social class."
cap22log <- "Probability of dying before age 90 for a 65-year-old by social class and period, women (top) and men (bottom). The last decades."
scap22log <- "Probability by social class, ages 65--99, last years."

##
vb3064 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) < 64.5, ]
vb3064$age <- factor(vb3064$age)
##vb3064$hisclass <- vb3064$hisclass40
```

```{r hazcalw}
source("R/getPHglm2.R")
source("R/getPHglm.R")
##nper <- 4
nper <- length(labb)
resw <- getPHglm2(vb3064[vb3064$sex == "female", ],
            nper, labb)
resm <- getPHglm2(vb3064[vb3064$sex == "male", ],
            nper, labb)
```

```{r cumsw, fig.cap = cap0, fig.scap = scap0, fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Women 30-64", ylim = c(0, 1))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, ylab = "H(65|40)",
                main = "Men 30-64", ylim = c(0, 1))##, exclude = 5:6)
par(oldpar)
```

@@

```{r cumswtab,cap="Table"}
vb3064$social_class <- vb3064$hisclass
vb3064$marital_status <- vb3064$civst
vb3064$urban_or_rural <- vb3064$urban
vb3064$age_group <- vb3064$age
fitt <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb3064, family = poisson)
dr <- drop1(fitt, test = "Chisq")
dr
##knitr::kable(dr)
##print(xtable(dr,label = "tab:cumswtab", caption = "Mortality: ANOVA, ages 40--64."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswstand, fig.cap = cap12, fig.scap = scap12, fig.height = 8, out.height = "14cm"}
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Women 30-64")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(65|40)", ylim = c(0.0, 0.6), main = "Men 30-64")##, exclude = 5:6)
par(oldpar)
```

@@

## Ages 65--99

```{r rawolddat,echo=FALSE}
rawskel <- get.file("rawskel")
rawskel$urban <- FALSE
rawskel <- rawskel[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
##
rawskume <- get.file("rawskume")
rawskume <- rawskume[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
##
rawfoli <- get.file("rawfoli")
rawfoli <- rawfoli[, c("id", "sex", "birthdate", "enter", "exit", "event", "period", "urban", "civst", "hisclass", "cohort")]
raw <- rbind(rawskel, rawskume, rawfoli)
raw <- age.window(raw, c(65, 100))
```

<!--
```{r getske6099}
vb6599 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
vb6599$age <- factor(vb6599$age) # NOTE:: Change to '6599'!
##vb6599$hisclass <- vb6599$hisclass60
```
-->
We now turn to the elderly, those who in most cases have left the workforce.
Corresponding to the previous analysis, we use the value of the cumulative hazards at
duration 25 (age 90), starting from age 65 as a summary measure. Looking at the table below, it is obvious that the statistical significance is still strong.

```{r cumswtabold,cap="Table"}
vb6599$marital_status <- vb6599$civst
vb6599$social_class <- vb6599$hisclass
vb6599$urban_or_rural <- vb6599$urban
vb6599$age_group <- vb6599$age
fittold <- glm(event ~ offset(log(exposure)) + urban_or_rural + marital_status + age_group + sex * social_class * period,
            data = vb6599, family = poisson)
drold <- drop1(fittold, test = "Chisq")
##knitr::kable(dr)
drold
##print(xtable(drold,label = "tab:cumswtabold", caption = "Mortality: ANOVA, ages 65--99."), 
  ##    type = "latex", comment = FALSE)
```

@@

```{r cumswold, fig.cap = cap1, fig.scap = scap1, fig.height = 8, out.height = "14cm"}
##source("R/plotED.R")
source("R/pgetPHglm2.R")
resw <- pgetPHglm2(vb6599[vb6599$sex == "female", ],
            nper, labb)
resm <- pgetPHglm2(vb6599[vb6599$sex == "male", ],
            nper, labb)

oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = FALSE,
                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Women 65-99")##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = FALSE,
                    ylab = "H(90|65)", ylim = c(0.0, 10.0), main = "Men 65-99")##, exclude = 5:6)
par(oldpar)
```

```{r cumsw2, fig.cap = cap22, fig.scap = scap22, fig.height = 8, out.height = "14cm"}
source("R/plotED.glm.R")
oldpar <- par(mfrow = c(2, 1))
woutp <- plotED.glm(resw, col = scol, logScale = TRUE, relative = TRUE, 
                    ylab = "H(90|65)",
                main = "Women 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
moutp <- plotED.glm(resm, col = scol, logScale = TRUE, relative = TRUE,
                    ylab = "H(90|65)",
                main = "Men 65-99", ylim = c(0, 0.6))##, exclude = 5:6)
par(oldpar)
```

```{r cumswoldtab,results='asis',eval=FALSE}
fitt <- glm(event ~ offset(log(exposure)) + urban + civst + age + sex * hisclass * period,
            data = vb6599, family = poisson)
dr <- drop1(fitt, test = "Chisq")
##knitr::kable(dr)
xtable(dr,label = "tab:cumswoldtab", caption = "ANOVA, all causes of death, 65-99.")
```
